### 2013-01-13 sections/evaluation.tex
caption consistency with graph legends
reduced the content to 7 pages

We used the first $20M$ records from a public flow trace to run a number of
flow queries using our benchmarking suite. We used the publically available
Trace 7 from the SimpleWeb \cite{simpleweb} repository for all the performance
evaluations. The input trace was compressed at \texttt{ZLIB\_LEVEL} $5$ using
the \texttt{zlib} suite. It was also converted to \texttt{nfdump} and SiLK
compatible formats and compressed with \texttt{zlib} keeping the same
compression level. The suite was run on a machine with $24$ cores of
$2.5$ GHz clock speed and $18$ GiB of memory.

The first set of queries attempt to stress the filter stage.  We use varying
values on the \texttt{packet} field offset to determine the amount of flow
records that are passed by the filter. The resultant filtered records are
written to \texttt{flow-tools} compatible file format and compressed at
\texttt{ZLIB\_LEVEL} $5$. The ratio of the number of filtered records in the
output trace to the number of the flow records in the input trace is plotted
against time. The evaluation results are shown in Fig.
\ref{fig:benchmarks-filter}.

@@

It can be seen that the performance of the filter stage in \ac{NFQL} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then again SilK and \texttt{nfdump} also use their own file
format. As a result, the amount of data that needs to be read (or written) may
be different to what it is for \ac{NFQL} and \texttt{flowtools}.
\texttt{nfdump} appears to be significantly faster than the rest. This is
because \texttt{nfdump} lacks \texttt{zlib} support, and as such the files
that are read and written use the \texttt{lzo} compression scheme. It is
important to note, that all the tools were single-threaded in this evaluation,
and did not completely utilize the $24$ available cores. It comes as a
realization, that adding \texttt{lzo} compression will drastically improve
\ac{NFQL}'s filter performance.

The second set of queries attempt to stress the grouper stage. We reuse the
filter query that produces a $1.0$ ratio to allow the grouper to receive the
entire trace as a filtered recordset. The grouper part of the query then
gradually increases the number of grouping terms in the \ac{DNF} expression to
increase the output/input ratio. The resultant groups are again written as
\texttt{flowtools} files using the same \texttt{zlib} compression level. The
ratio of the number of groups formed to the number of the input filtered
records is plotted against time. \texttt{nfdump} and \texttt{flowtools} do not
support grouping, and therefore are not considered in this evaluation. The
results are shown in Fig. \ref{fig:benchmarks-grouper}.

@@

The evaluation graph reveals that the performance of the \ac{NFQL} grouper
stage is close to the performance of SiLK. The time taken by the tools are
comparable on lower ratios, but on higher ratios, \ac{NFQL} starts to drift
apart. Since most of the time is taken in writing the records to files, it is
unclear whether SiLK's usage of its own file format, which may reduce
reads/writes, is responsible for the drift on higher ratios. SiLK's
\texttt{rwgroup} tool is also supplied a \texttt{-{}-summarize} flag to force it
to write only the first record of each group, to make both tools write the
same number of records to files. This gives SiLK the leverage to not store
information about which members are part of the group. \ac{NFQL} on the other
hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples. It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took an advantage of an inherent concurency arising from how
the query is structured as one single \texttt{bash} script using pipes. The
pipe between \texttt{rwsort} and \texttt{rwgroup} makes the two processes run
concurrently, the effect of which gets more pronounced on higher ratios and
can be a drift determining factor.  The profiling results from GNU
\texttt{gprof} \cite{graham:1982} indicate that $60\%$ of the time is taken in
\texttt{qsort} comparator calls.  As a result, it comes as no surprise, that
bifurcating \texttt{qsort} invocation to multiple threads and later merging
the results back using merge sort will help parallelize the grouper stage and
maybe reduce the drift on higher ratios. In addition, since all of the
evaluation queries had grouping terms using an equality comparator, \ac{NFQL}
can introspect such a grouping rule to dynamically optimize processing
searches using a hashtable and turn to \texttt{qsort} based grouping only as a
fallback.

The third set of queries attempt to stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the \texttt{packet} field offset
to determine the amount of groups that are filtered ahead. The filtered groups
are again written as \texttt{flowtools} files using the same \texttt{zlib}
compression level. The ratio of the number of output filtered groups to the
number of the input group is plotted against time. The results are shown in
Fig. \ref{fig:benchmarks-groupfilter}.

@@

The fourth set of queries attempt to stress the merger stage. We reuse the
filter, grouper and group filter queries that produce a $1.0$ ratio. These
queries are then run in two separate branches to produce identical filtered
group records.  The merger then applies match rules to produce different
output to input ratios. The groups that are merged are again written as
\texttt{flowtools} files using the same \texttt{zlib} compression level. The
ratio of the number of merged groups to twice\footnote{Each branch pushes the
entire trace as an input to the merger.} the number of flow records in the
original trace is plotted against time. A data point for SiLK for the $0.2$
ratio is not available since the \ac{NFQL} query executed at that data point
uses \texttt{OR} expressions, which are not supported by SiLK. As a result, an
equivalent SiLK query is not formulated.

\begin{figure}[ht!]
  \begin{center}
    \includegraphics* [width=0.9\linewidth]{merger}
    \caption{Merger Stage}
    \label{fig:benchmarks-merger}
  \end{center}
\end{figure}

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flowtools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \ac{NFQL} merger implementation. The optimized merger
implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also writes
a merged group record to a file only once despite the number of times it has
matched. Without these optimizations, running such queries on the merger would
keep the CPU churning for days without results.

The last set of queries attempt to stress the ungrouper stage. They reuse the
entire merger queries as is, but enable the ungrouper as well. This means,
that the ungrouper now attempts to unfold the merged groups returned by the
merger to write their member flow records to \texttt{flowtools} files.
However, since the merger receives each flow record as its own filtered group,
each merged group has only one member. As a result, the ungrouper ends up
rewriting the merged groups as \texttt{flowtools} files using the same
\texttt{zlib} compression level. This means that the execution engine end up
taking twice the amount of time than the merger. It is important to note, that
SiLK does not have such an equivalent ungrouping tool and is therefore not
considered in this final evaluation.

>>>>

We used the first $20M$ records from the publically available Trace 7 from the
SimpleWeb \cite{simpleweb} repository for our performance evaluations. The
input trace was compressed at \texttt{ZLIB\_LEVEL} $5$ using the \texttt{zlib}
suite. It was also converted to \texttt{nfdump} and SiLK compatible formats
keeping the same compression level. The suite was run on a machine with $24$
cores of $2.5$ GHz clock speed and $18$ GiB of memory.

The first set of queries stress the filter stage.  We use varying values on
the \texttt{packet} field offset to control the amount of flow records that
are passed by the filter. The resultant filtered records are written as
\texttt{flow-tools} files and compressed at \texttt{ZLIB\_LEVEL} $5$. The
ratio of the number of filtered records in the output trace to the number of
the flow records in the input trace is plotted against time. The results are
shown in Fig. \ref{fig:benchmarks-filter}.

@@

It can be seen that the performance of the filter stage in \ac{NFQL} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then SilK and \texttt{nfdump} also use their own file
format. \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} uses the \texttt{lzo} compression scheme. It comes
as a realization, that adding \texttt{lzo} compression will drastically
improve \ac{NFQL}'s filter performance. It is important to note, that all the
tools were single-threaded in this evaluation, and did not completely utilize
the $24$ available cores.

The second set of queries stress the grouper stage. We reuse the filter query
that produces a $1.0$ ratio to allow the grouper to receive the entire trace
as a filtered recordset. The grouper part of the query then gradually
increases the number of grouping terms in the \ac{DNF} expression to increase
the output/input ratio. The resultant groups are again written as
\texttt{flowtools} files using the same \texttt{zlib} compression level. The
ratio of the number of groups formed to the number of the input filtered
records is plotted against time. \texttt{nfdump} and \texttt{flowtools} do not
support grouping, and therefore are not considered in this evaluation. The
results are shown in Fig. \ref{fig:benchmarks-grouper}.

@@

The evaluation graph reveals that the time taken by the tools are comparable
on lower ratios, but on higher ratios, \ac{NFQL} starts to drift apart. Since
most of the time is taken in writing the records to files, it is unclear
whether SiLK's usage of its own file format is responsible for the drift.
SiLK's \texttt{rwgroup} tool is also supplied a \texttt{-{}-summarize} flag to
force it to write only the first record of each group, to make both tools
write the same number of records. This gives SiLK the leverage to not store
information about which members are part of the group. \ac{NFQL} on the other
hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples.  It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took advantage of an inherent concurency arising from a pipe
between \texttt{rwsort} and \texttt{rwgroup}, which makes the two processes
run concurrently, the effect of which gets more pronounced on higher ratios.
The profiling results from GNU \texttt{gprof} \cite{graham:1982} indicate that
$60\%$ of the time is taken in \texttt{qsort} comparator calls.  As a result,
it comes as no surprise, that bifurcating \texttt{qsort} invocation to
multiple threads and later merging the results back using merge sort will help
parallelize the grouper stage and maybe reduce the drift on higher ratios. In
addition, since all of the evaluation queries had grouping terms using an
equality comparator, \ac{NFQL} can introspect such a grouping rule to
dynamically optimize processing searches using a hashtable and turn to
\texttt{qsort} based grouping only as a fallback.

The third set of queries stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the \texttt{packet} field offset
to control the amount of groups that are filtered ahead. The filtered groups
are again written as \texttt{flowtools} files using the same \texttt{zlib}
compression level. The ratio of the number of output filtered groups to the
number of the input group is plotted against time. The results are shown in
Fig. \ref{fig:benchmarks-groupfilter}.

@@

The fourth set of queries stress the merger stage. We reuse the filter,
grouper and group filter queries that produce a $1.0$ ratio. These queries are
then run in two separate branches to produce identical filtered group records.
The merger then applies match rules to produce different output to input
ratios. The groups that are merged are again written as \texttt{flowtools}
files using the same \texttt{zlib} compression level. The ratio of the number
of merged groups to twice\footnote{Each branch pushes the entire trace as an
input to the merger.} the number of flow records in the original trace is
plotted against time. A data point for SiLK for the $0.2$ ratio is not
available since the \ac{NFQL} query executed at that data point uses
\texttt{OR} expressions, which are not supported by SiLK.

\begin{figure}[ht!]
  \begin{center}
    \includegraphics* [width=0.9\linewidth]{merger}
    \caption{Merger Stage: NFQL vs SiLK}
    \label{fig:benchmarks-merger}
  \end{center}
\end{figure}

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flowtools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \ac{NFQL} merger implementation. The optimized merger
implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also
writes a merged group record to a file only once despite the number of times
it has matched. Without these optimizations, running such queries on the
merger kept the CPU churning for days without results.

The last set of queries stress the ungrouper stage. They reuse the entire
merger queries as is, but enable the ungrouper as well. This means, that the
ungrouper now attempts to unfold the merged groups returned by the merger to
write their member flow records to \texttt{flowtools} files.  However, since
the merger receives each flow record as its own filtered group, each merged
group has only one member. As a result, the ungrouper ends up rewriting the
merged groups as \texttt{flowtools} files using the same \texttt{zlib}
compression level. This means that the execution engine end up taking twice
the amount of time than the merger. It is important to note, that SiLK does
not have such an equivalent ungrouping tool and is therefore not considered in
this final evaluation.

### 2013-01-13 sections/implementation/adaptable-compression-levels.tex
reduced the content to 7 pages

The flow-records echoed to the standard output can also be written to Netflow
$v5$ \texttt{flow-tools} files. The \texttt{-{}-dirpath} switch allows one to
provide a directory path where the results can be stored. Results from each
stage of the pipeline can also be written to separate files with the increase
in the verbosity level. In fact, \texttt{-{}-dirpath} and \texttt{-{}-verbose}
work well together to adjust the level at which the writes are to be made.

@@

The engine uses the \texttt{zlib} \cite{rfc1950} software library to compress
the results written to the \texttt{flow-tools} files. \texttt{zlib} supports
$9$ compression levels with $9$ being the highest. The \ac{NFQL} engine
supports \texttt{-{}-zlevel} switch to allow the user to supply its desired
choice of the of the compression level. A default level of $5$ is used for
writes if the switch is not supplied during runtime.  Fig.
\ref{fig:engine-zlevel} shows the time taken to write a sample of records
passing the filter stage for each \texttt{z-level}. It goes to show that each
level adds its own performance overhead and must be used with discretion. It
is also important to note that other tools may use different compression
algorithms. \texttt{nfdump} for instance, uses \texttt{lzo} \cite{lzo}
compression to trade space for faster compression and decompression.

>>>>

@@

The engine uses the \texttt{zlib} \cite{rfc1950} software library to compress
the results written to the \texttt{flow-tools} files. \texttt{zlib} supports
$9$ compression levels with $9$ being the highest. \ac{NFQL} allows the user
to supply its desired choice of the compression level.  A default level of $5$
is used for writes if a choice is not indicated.  Fig.
\ref{fig:engine-zlevel} shows the time taken to write a sample of records
passing the filter stage for each \texttt{z-level}. It goes to show that each
level adds its own performance overhead and must be used with discretion. It
is also important to note that other tools may use different compression
algorithms.  \texttt{nfdump} for instance, uses \texttt{lzo} \cite{lzo}
compression to trade space for faster compression and decompression.

### 2013-01-13 sections/implementation/execution-workflow.tex
reduced the content to 7 pages

In order to be able to make comparisons on the field offsets of such a term,
the comparator needs to know the type of the comparison and the length of the
field offset. This information is parsed by execution engine once the query is
read and is therefore not available at compile time.  In order to subvert the
need to define complex branching statements, a dedicated comparator is defined
for every possible field length and comparison operation. A Python script
generates C source code for these comparators at compile time conforming to
the structure shown in listing \ref{lst:filter-struct}. This allows the term
definitions to make runtime calls using a function name derived from the
combination of operation type and field length.

@@

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This eliminated the need to copy all the
flow-records when moving ahead in the processing pipeline. As a result, there
is no dedicated splitter stage in the execution engine. Each branch references
the flow records from a common memory location. This helps keep the memory
costs at a minimum when multiple branches are involved.

@@

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. The filtered records are saved in a common location from
where they are referenced by each branch. This helps keep the memory costs at
a minimum when multiple branches are involved.  We used the publicly
available Trace 7 from the SimpleWeb \cite{simpleweb} to compare the
performance of the optimized \ac{NFQL} implementation against the one defined
by the flow query language as shown in Fig.  \ref{fig:fv1-fv2-filter}. The
filter stage implementation with these optimizations runs 10 times faster and
is more pronounced on higher ratios.

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on the field offsets all the requested
grouping terms of a clause. This helps the execution engine perform a nested
binary search to reduce the linear pass to a fairly small filtered recordset.
However, the number of elements upon which the search has to be performed
needs to be known at each level of the binary lookup. In order to avoid a
linear run to count the number of elements, the parent level binary search
invocation returns a boundary with the first and last element to enable such a
calculation. This helps the grouper perform faster search lookups to find
records that must group together in $O(n*lg(k))$ time with a preprocessing
step taking $O(p*n*lg(n))$ in the average case, where $n$ is the number of
filtered records, $p$ is the number of grouping terms in a clause and $k$ is
the number of unique filtered records.

The grouping approach has further been optimized when the filtered records are
grouped for equality. In such a scenario, the need to search for unique
records and a  subsequent binary search goes away.  The groups can now be
formed in $O(n)$ time with the same preprocessing step taking $O(p*n*lg(n))$.
The performance evaluation of the grouper handling this special case against
its behavior when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}. Profiling the performance of the grouper on queries
that produce higher ratios reveal that a significant amount of time is taken
in binary search which is eliminated in the special case of equality
comparisons.

@@

\subsubsection{Group Filter} The groupfilter stage is used to filter the
groups produced by the grouper based on some absolute rules defined in a
\ac{DNF} expression. The \texttt{struct} term holds information about the flow
record offset, the value being compared to and the operator which maps to a
unique \texttt{enum} value.  This enum value is used to map the operation to a
specific group-filter function. The group-filter functions are auto-generated
using a Python script.

@@

The merger is used to relate groups from different branches according to a
merging criterion. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
runtime. As a result, an iterator that can provide all possible permutations
of $m-$tuple (where $m$ is the number of branches) group record $ID$s was
needed.  The result of the iterator can then be used to make a match. The
merger stage begins by initializing this iterator passing it the number of
branches, and information about each branch. Then, it loops over to get a new
$m-$tuple of group record $ID$s on each iteration until the iterator returns
\texttt{false}.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. The \ac{NFQL} implementation optimizes the
merger to skip over iterator permutations when a state of a current field
offset value may not allow any further match beyond the index in the current
branch.  For such an optimization to work, the filtered group records must be
sorted in the order of field offsets specified in the merger clause.
Specifying the filtered group records in any other order may lead to undefined
behavior. This means, that if the same field offsets were used in the grouper
stage, the terms in the group clause can be rearranged by the query designer
to align with the order of terms in the merger clause.

The flow query language also bases the merger matches on the notion of matched
tuples. This means that a filtered group record can be written to a file
multiple times if it is part of multiple matched tuples. This situation is
very common and it worsens when different branches have similar filtered
groups records. Since, the function of the merger is to find a match of groups
records across branches based on a predefined condition, all the group records
across branches that satisfy the condition can be clubbed into one collection
instead of separate tuples. All the group records within a collection can then
be written to the file. This eliminates the inherent redundancy and
significantly improves the merger performance.

The performance comparison of this approach against the one suggested by the
specification is tricky. The merger implementation of the original
specification is slow. It is so slow that it keeps churning the CPU for days
without results. The newer approach takes less than an hour. The performance
evaluation of this newer approach is discussed in more details in the next
section

>>>>

In order to be able to make comparisons on the field offsets of a term, the
comparator needs to know the type of the comparison and the length of the
field offset. This information is parsed by execution engine once the query is
read and is therefore not available at compile time.  In order to subvert the
need to define complex branching statements, a dedicated comparator is defined
for every possible field length and comparison operation. A Python script
generates C source code for these comparators at compile time conforming to
the structure shown in listing \ref{lst:filter-struct}. This allows the term
definitions to make runtime calls using a function name derived from the
combination of operation type and field length.

@@

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This eliminated the need to copy all the
flow-records when moving ahead in the pipeline. As a result, there is no
dedicated splitter stage in the execution engine. Each branch references the
flow records from a common memory location. This helps keep the memory costs
at a minimum when multiple branches are involved.

@@

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publicly available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of \ac{NFQL} against the
one defined by the flow query language as shown in Fig.
\ref{fig:fv1-fv2-filter}. The filter stage implementation with these
optimizations runs 10 times faster and is more pronounced on higher ratios.

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on the field offsets of all the requested
grouping terms of a clause. This helps the execution engine perform a nested
binary search to reduce the linear pass to a fairly small filtered recordset.
This helps the grouper perform faster search lookups to find records that must
group together in $O(n*lg(k))$ time with a preprocessing step taking
$O(p*n*lg(n))$ in the average case, where $n$ is the number of filtered
records, $p$ is the number of grouping terms in a clause and $k$ is the number
of unique filtered records. The grouping approach has further been optimized
when the filtered records are grouped for equality. In such a scenario, the
need to search for unique records and a  subsequent binary search goes away.
The groups can now be formed in $O(n)$ time with the same preprocessing step
taking $O(p*n*lg(n))$.  The performance evaluation of the grouper handling
this special case against when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}.

@@

\subsubsection{Group Filter} The groupfilter stage is used to filter the
groups based on some absolute rules defined in a \ac{DNF} expression. The
\texttt{struct} term holds information about the flow record offset, the value
being compared to and the operator which maps to a unique \texttt{enum} value.
This enum value is used to map the operation to a specific group-filter
function.

@@

The merger is used to relate groups from different branches according to a
merging criterion. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
runtime. As a result, an iterator that can provide all possible permutations
of $m-$tuple (where $m$ is the number of branches) group record $ID$s was
implemented. The result of the iterator is then be used to make a match.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. \ac{NFQL} optimizes the merger to skip over
iterator permutations when a state of a current field offset value may not
allow any further match beyond the index in the current branch.  For such an
optimization to work, the filtered group records must be sorted in the order
of field offsets specified in the merger clause.  Specifying the filtered
group records in any other order may lead to undefined behavior. This means,
that if the same field offsets were used in the grouper stage, the terms in
the group clause can be rearranged by the query designer to align with the
order of terms in the merger clause.

The flow query language also bases the merger matches on the notion of matched
tuples. This means that a filtered group record can be written to a file
multiple times if it is part of multiple matched tuples. This situation
worsens when different branches have similar filtered groups records. Since,
the function of the merger is to find a match of groups records across
branches based on a predefined condition, all the group records across
branches that satisfy the condition can be clubbed into one collection instead
of separate tuples. All the group records within a collection can then be
written to the file. This eliminates the inherent redundancy and significantly
improves the merger performance.

### 2013-01-13 sections/implementation/flow-query-intermediate-format.tex
reduced the content to 7 pages

Each stage of the processing pipeline is expressed in the JSON query as a
\ac{DNF} expression. The elements of the conjuctive clauses are terms.
\texttt{json-c} \cite{jsonc} is used to parse the flow query file. The
pipeline stages of the flow query language are also encapsulated in Python
classes and scripts have been written to facilitate the front-end parser to
serialize these pipeline objects to a JSON flow query.  The mapping
of the query to the structs defined in the execution engine is not trivial.
When reading the JSON query at runtime, the field names of a NetFlow
$v5$ record are read in as strings.  Utility functions are defined that map
the field names to internal struct offsets and the field types and the
operations to internal unique enum members.

>>>>

Each stage of the pipeline is expressed in the JSON query as a \ac{DNF}
expression. The elements of the conjuctive clauses are terms.  \texttt{json-c}
\cite{jsonc} is used to parse the flow query file. The mapping of the query to
the structs defined in the execution engine is not trivial.  When reading the
JSON query at runtime, the field names of a NetFlow $v5$ record are read in as
strings.  Utility functions are defined that map the field names to internal
\texttt{struct} offsets and the field types and the operations to internal
\texttt{enum} members.

### 2013-01-13 sections/implementation/performance-optimizations.tex
reduced the content to 7 pages

\ac{NFQL} has dedicated comparator functions for each type of operation and
the type of the field offset it operates upon. It is not guaranteed that given
the type of the query and trace, the program will eventually go through each
stage of the pipeline. It is also possible that the program exits before,
because there is nothing more for the next stage to compute. The function
pointers are therefore set as late as possible and are called from their
respective stages just before the comparison is needed.  As a result, we save
the computation time wasted in setting the function pointer for stage if it is
is never executed.

\ac{NFQL} pushes the filter stage out of the branches and closer to where the
trace is originally read. This reduces the memory allocation to only records
that passed the filter stage, thereby reducing the memory footprint of the
execution engine.

Each stage of the processing pipeline is dependent on the result of the
previous one. As a result, the stages should only proceed, when the previous
returned results. Implementing such a response was straightforward for the
grouper and group filter, the merger although was a little trickier.  The
merger stage proceeds only when every branch has non-zero filtered groups.
The iterator initializer \texttt{iter\_init(\ldots)} deallocates and returns
\texttt{NULL} if any one branch has $0$ filtered groups.  Consequently a check
is performed in the merger to make sure \texttt{iter} is \emph{not}
\texttt{NULL}.

The results from each stage of the pipeline are echoed to the standard output
just before the execution engine exits. This leads to an additional loop to
echo the results, however echoes to the standard output are only used for
debugging purposes. The writes to a file can, however, be legitimately
requested in a real network analysis task. As a result, it is essential to
avoid additional loops when performing writes to a file. The execution engine
therefore writes each result record to a file as soon as it is seen by the
pipeline stage.

>>>>

\ac{NFQL} has dedicated comparator functions for each type of operation and
the type of the field offset it operates upon. It is not guaranteed that given
the type of the query and trace, the program will eventually go through each
stage of the pipeline. It is also possible that the program exits before,
because there is nothing more for the next stage to compute. The function
pointers are therefore set as late as possible and are called from their
respective stages just before the comparison is needed.

Each stage of the processing pipeline is dependent on the result of the
previous one. As a result, the stages should only proceed, when the previous
returned results. Implementing such a response was straightforward for the
grouper and group filter, the merger although was a little trickier.  The
merger stage proceeds only when every branch has non-zero filtered groups.
The iterator initializer deallocates and returns \texttt{NULL} if any one
branch has $0$ filtered groups.  Consequently a check is performed in the
merger to make sure \texttt{iter} is \emph{not} \texttt{NULL}.

The flow-records echoed to the standard output can also be written as
\texttt{flow-tools} files. In fact, results from each stage of the pipeline
can be written to separate files with the increase in the verbosity level.
This leads to additional loops over the resultsets if the writes are made at
the end of the processing pipeline. The execution engine therefore writes each
result record to a file as soon as it is seen by the pipeline stage.

### 2013-01-13 sections/introduction.tex
reduced the content to 7 pages

Researchers, service providers and security analysts have long been interested
in network and user behavioral patterns of the traffic crossing the internet
backbone. They want to use this information for the purpose of billing and
mediation, bandwidth provisioning, detecting malicious attacks, network
performance evaluation and overall improvement. Traffic measurement techniques
that have been rapidly evolving in the last decade have today matured enough
to provide such an insight.

Capturing network flow records has emerged out to be one of the favored
network measurement techniques. In this technique, packets traversing an
obvervation point are not captured raw, instead common characteristics of the
packet header are aggregated into flow records. They are then exported to a
collector for further analysis.  This reduces the amount of traffic aggregated
at the collector.

NetFlow and \ac{IPFIX} are the two popular protocols for IP flow information
export. NetFlow \cite{rfc3954} is a network protocol designed by Cisco
Systems, which allows routers to generate and export flow records to a
designated collector. NetFlow v$9$ provides flexibility through user-tailored
export templates, \ac{MPLS} and IPv$6$ support and a larger set of flow keys.
\ac{IPFIX} \cite{rfc5101} on the other hand is an open standard by IETF based
on NetFlow v$9$. The novelty of the standard lies in its ability to describe
record formats at runtime using templates based on an extensible and
well-defined information model. The data transfer mechanism is also simplistic
and extensible by being unidirectional and transport protocol agnostic.

>>>>

Researchers, service providers and security analysts have long been interested
in network and user behavioral patterns of the traffic crossing the internet
backbone. Traffic measurement techniques that have been rapidly evolving in
the last decade have today matured enough to provide insights based on these
traffic patterns. Capturing network flow records is one of the favored network
measurement technique used for this task. In this technique, packets
traversing an obvervation point are not captured raw, instead common
characteristics of the packet header are aggregated into flow records. They
are then exported to a collector for further analysis.  This reduces the
amount of traffic aggregated at the collector.

NetFlow and \ac{IPFIX} are the two popular protocols for IP flow information
export. NetFlow \cite{rfc3954} is a network protocol designed by Cisco
Systems, which allows routers to generate and export flow records to a
designated collector. \ac{IPFIX} \cite{rfc5101} on the other hand is an open
standard by IETF based on NetFlow v$9$. The novelty of the standard lies in
its ability to describe record formats at runtime using templates based on an
extensible and well-defined information model. The data transfer mechanism is
also simplistic and extensible by being unidirectional and transport protocol
agnostic.
